Very simplified likelihood

### 2024-01-05 inst/very_simplified_likelihood.Rmd
distinguish prob from likelihood

Consider a race between a competitor of Bradley-Terry strength $a$ and
$n$ competitors each of strength $b$; we require $a+b=1$.  An
observation is indexed by $r$, the number of $b$ clones finishing
ahead of $a$, so $0\leqslant r\leqslant n$.  The initial field
strength is $a+nb=n+(n-1)a$.

@@

where $B=1/b$.

@@

Placing last is finishing  after $n$ clones finish:

@@

\[
n=3\longrightarrow\mathcal{L}=\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^3}{(1+b)(1+2b)}
\]

\[
n=4\longrightarrow\mathcal{L}=\frac{b}{a+4b}\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=
\frac{b^4}{(1+b)(1+2b)(1+3b)}
\]

@@

\[
\frac{1}{B(B+1)(B+2)(B+3)}=\frac{(B-1)!}{(B+3)!}=\frac{\Gamma(B)}{\Gamma(B+4)}
\]

\[
\frac{B-1}{(B+n-1)(B+n-2)\cdots(B+1)\cdot B\cdot(B-1)}
\]

@@

```{r}
f_single <- function(a,r,n,log=FALSE){  # not vectorised
  B <- 1/(1-a)
  if(log){
     if(r==n){
       out <- lgamma(B)-lgamma(B+n)
     } else {
       out <- log(B-1) - sum(log(B + B+(n-r-1):(n-1)))
     }
   } else {  # not-log 
     if(r==n){
      out <- gamma(B)/gamma(B+n)
     } else {
       out <- (B-1)/prod(B+(n-r-1):(n-1))
     }
   }	
   return(out)
}x

@@

Now some visuals.  Suppose $r=4,n=8$:

>>>>

Consider a race between a competitor of Bradley-Terry strength $a$ and
$n$ competitors each of strength $b$; we require $a+b=1$.  NB we have
$n+1$ competitors in total, 1 of strength $a$ and $n$ of strength $b$.
An observation is indexed by $r$, the number of $b$ clones finishing
ahead of $a$, so $0\leqslant r\leqslant n$.  The initial field
strength is $a+nb=n+(n-1)a$.

@@

where $B=1/b$ [also using the fact that $a+b=1$].

@@

Placing last is finishing  after $n=r$ clones finish:

@@

\[
n=3\longrightarrow\mathcal{L}=\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^3}{(1+2b)(1+b)}
\]

\[
n=4\longrightarrow\mathcal{L}=\frac{b}{a+4b}\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=
\frac{b^4}{(1+3b)(1+2b)(1+b)}
\]

@@

\[
\frac{1}{(B+3)(B+2)(B+1)B}=\frac{(B-1)!}{(B+3)!}=\frac{\Gamma(B)}{\Gamma(B+4)}
\]

In general we would have

\[
\frac{1}{(B+n)(B+n-1)\cdots(B+1)B}=\frac{(B-1)!}{(B+n)!}=\frac{\Gamma(B)}{\Gamma(B+n+1)}
\]

@@

```{r}
f_single <- function(a,r,n,log=FALSE){  # not vectorised
  B <- 1/(1-a)
  if(log){
     if(r==n){  # last place
       out <- lgamma(B)-lgamma(B+n)
     } else {   # not last
       out <- log(B-1) - sum(log(B + B+(n-r-1):(n-1)))
     }
   } else {  # not-log 
     if(r==n){
      out <- gamma(B)/gamma(B+n)
     } else {
       out <- (B-1)/prod(B+(n-r-1):(n-1))
     }
   }	
   return(out)
}

@@

## Note on the differences between probability and likelihood

The likelihood calculations above can be confusing becuase likelihood
is not the same as probability.

Consider, for example, $n=2$ so we have 3 competitors.

probability of focal competitor coming first [i.e. $r=0$]:

\[\frac{a}{a+2b}=\frac{1-b}{1+b}\]

Now what is the probability of coming second, that is, $r=1$?

My first thought was:

\[
\mbox{prob of coming second}   =
\frac{b}{a+2b}\cdot\frac{a}{a+b}=\frac{b(1-b)}{1+b}\qquad\mbox{WRONG}
\]

but this is incorrect: there are two clones of (strength) $b$, and the
likelihood arguments above do not distinguish between their finishing
order.  We can name the competitors $a$, $b_1$ and $b_2$ and specify
that $b_1$ and $b_2$ both have strength $b$.  Then "coming second"
means that the finishing order was $b_1,a,b_2$ or $b_2,a,b_1$.  Noting
that these events are disjoint, the probability would be

\[\mbox{prob of coming second}   =
\operatorname{P}(b_1,a,b_2)      +
\operatorname{P}(b_2,a,b_1)      =
\frac{b}{a+2b}\cdot\frac{a}{a+b} +
\frac{b}{a+2b}\cdot\frac{a}{a+b} = \frac{2(1-b)}{1+b}
\]

Similarly for the focal competitor coming last we need to sum the
probabilities of finishing orders $b_1,b_2,a$ and $b_2,b_1,a$:

\[\mbox{prob of coming third}=
\operatorname{P}(b_1,b_2,a) +
\operatorname{P}(b_2,b_1,a) =
\frac{b}{a+2b}\cdot\frac{b}{a+b}+
\frac{b}{a+2b}\cdot\frac{b}{a+b}=
\frac{2b^2}{1+b}.
\]

Above we see a factor of two difference between the probabilities and
the likelihoods presented previously.  Note that there is no such
factor for the probability of the focal competitor finishing first:

\[
\mbox{prob of coming first}      =
\operatorname{P}(a,b_1,b_2)      +
\operatorname{P}(a,b_2,b_1)      =
\frac{b}{a+2b}\cdot\frac{b}{b+b} +
\frac{b}{a+2b}\cdot\frac{b}{b+b} = \frac{1-b}{1+b}
\]

Now we see that the probabilities sum to

\[
\frac{1-b}{1+b} + \frac{2b^2}{1+b} + \frac{2(1-b)}{1+b}=1
\]

Now try $r=3$, 4 competitors.   Taking things one step at a time:

\[
\mbox{prob of coming first}      =
3!\cdot\frac{a}{a+3b}\cdot\frac{b}{3b}\cdot\frac{b}{2b}\cdot\frac{b}{b}
\]

## Visuals

Now some likelihood visuals.  Suppose $r=4,n=8$:
